An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-02-18 EconGrowth.Rnw
Added some CES information to the US and UK Excel workbooks. Added a footnote about Stern and Kander's equation being wrong in the text of their 2012 paper.

\usepackage{soul} %Provides strikethrough text
\usepackage{float} %Allows precise positioning of tables and figures within the text

@@

<<setup, echo=FALSE, message=FALSE>>=
require(knitr)
require(lattice)
require(latticeExtra)
require(ggplot2)
require(car)
require(mosaic)
require(xtable)
require(nlmrt)
# tikz allows use of LaTeX formatting and font in graphs. Allows for a consistent look across the paper.
# See http://r-forge.r-project.org/R/?group_id=440 for instructions on installing tikzDevice.
require(tikzDevice) 
trellis.par.set(theme=theme.mosaic())
options(width=75)
options(show.signif.stars=TRUE)
opts_chunk$set(
  # ********************************************************************************************************
  eval=FALSE,
  # ********************************************************************************************************
  warning=FALSE, #Tells whether to show warnings in the output.
  dev='tikz',    #Allows LaTeX code in graphical output. E.g., "$y$" for a variable name in the legend for a graph.
  tidy=FALSE,
  comment=NA,
  #Tells whether to cache output from chunks are saved. Cacheing saves time. 
  #However, references to LaTeX tables DO NOT WORK if cacheing is turned on (cache=TRUE).
  cache=FALSE,
  echo=FALSE,    #Tells whether to echo code for all chunks. Set TRUE on an individual chunk to see its code.
  message=FALSE) #Turns off messages for all chunks. Set TRUE on an individual chunk to see it.

@@

@@

<<single-factor functions, eval=TRUE>>=
####################
# Returns an nls single-factor model for the country and factor specified.
# factor should be one of "K", "L", "Q", "X", or "U".
##
singleFactorModel <- function(countryAbbrev, factor){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  # We'll change the name of the desired column to "f"
  dataTable <- replaceColName(dataTable, factor, "f")
  # Now do the fit.
  lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
  # Some economies need different guesses
  if (countryAbbrev == "IR" && factor == "X"){
    mGuess <- 0.7
  } else {
    # This works for every other combination of country and factor.
    mGuess <- 0.5
  }
  # Runs a non-linear least squares fit to the data. We've replaced beta with 1-alpha for simplicity.
  modelSF <- nls(iGDP ~ exp(lambda*iYear) * f^m, 
                 data=dataTable, 
                 start = list(lambda=lambdaGuess, m=mGuess)
                 )
  return(modelSF)
}

@@

\noindent to historical data where the factor of production ($f$) is any of $k$, $l$, $q$, $x$, or $u$, and the exponent ($m$) is $\alpha$ for $k$, $\beta$ for $l$, or $\gamma$ for $q$, $x$, or $u$.

@@

<<cobb-douglas functions, eval=TRUE>>=
####################
# Returns an nls Cobb-Douglas model for the country specified
# Give an energyType ("Q", "X", or "U") if you want to include an energy term
##
cobbDouglasModel <- function(countryAbbrev, energyType){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  #Run the non-linear least squares fit to the data based on whether we have an energy type or not.
  if (missing(energyType)){
    # No energy term desired in the Cobb-Douglas equation.
    # Establish guess values for alpha and lambda.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.7 #0.7 gives good results for all countries.  
    # Runs a non-linear least squares fit to the data. We've replaced beta with 1-alpha for simplicity.
    modelCD <- nls(iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^(1 - alpha), 
                   data=dataTable, 
                   start = list(lambda=lambdaGuess, alpha=alphaGuess),
                   #Include the next 3 lines to fit with constraints.
                   #algorithm = "port",
                   #lower = list(lambda=-Inf, alpha=0), #constraints on lambda and alpha
                   #upper = list(lambda=Inf, alpha=1)
                   )
    return(modelCD)
  } else {
    # We need to do the Cobb-Douglas fit with the desired energy data.
    # To achieve the correct fit, we'll change the name of the desired column
    # to "iEToFit" and use "iEToFit" in the nls function. 
    dataTable <- replaceColName(dataTable, energyType, "iEToFit")
    # Reparameterize to ensure that we meet the constraints:
    # * alpha + beta + gamma = 1.0.
    # * alpha, beta, and gamma are all between 0.0 and 1.0.
    # To do this, we reparameterize as
    # * 0 < a < 1
    # * 0 < b < 1
    # * alpha = min(a, b)
    # * beta = b - a
    # * gamma = 1 - max(a, b)
    # Establish guess values for lambda, alpha, and beta.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.9
    betaGuess <- 1.0 - alphaGuess
    aGuess <- alphaGuess
    bGuess <- alphaGuess + betaGuess
    # Now actually do the fit, using the column name "iEToFit".
    modelCDe <- nls(iGDP ~ exp(lambda*iYear) * iCapStk^min(a,b) * iLabor^abs(b-a) * iEToFit^(1.0 - max(a,b)), 
                    data = dataTable,
                    start = list(lambda=lambdaGuess, a=aGuess, b=bGuess),
                    control = nls.control(maxiter = 200, 
                                          tol = 1e-05, 
                                          minFactor = 1/1024, 
                                          printEval=FALSE, #Tells whether to print details of curve fit process.
                                          warnOnly=TRUE),
                    #Include the next 3 lines to fit with constraints.
                    #algorithm = "port",
                    #lower = list(lambda=-Inf, a=0, b=0),
                    #upper = list(lambda= Inf, a=1, b=1)
                    )
    return(modelCDe)
  }
}

@@

<<CES functions, eval=TRUE>>=
####################
# Returns an nls ces model for the country and energyType specified.
# energyType should be one of Q", "X", or "U".
##
cesModel <- function(countryAbbrev, energyType){
  # Load the data that we need.
  dataTable <- loadData(countryAbbrev)
  # We need to do the CES fit with the desired energyType.
  # To achieve the correct fit, we'll change the name of the desired column
  # to "iEToFit" and use "iEToFit" in the nls function.
  dataTable <- replaceColName(dataTable, energyType, "iEToFit")
  # Establish default guesses for sigma, gamma_E, beta, lambda_L and lambda_E.
  sigmaGuess <- 0.663 # 0.663 for Sweden
  gamma_EGuess <- 0.01 # Minimum contribution by energy is the default.
  betaGuess <- 0.7 # 0.7 for Sweden
  lambda_LGuess <- 0.011 # 0.011 for Sweden
  lambda_EGuess <- 0.018 # 0.018 for Sweden
  lower = list(sigma=-Inf, gamma_E=0.0, beta=-Inf, lambda_L=-Inf, lambda_E=-Inf)
  upper = list(sigma= Inf, gamma_E=1.0, beta= Inf, lambda_L= Inf, lambda_E= Inf)
  if (countryAbbrev == "US"){
    gamma_EGuess <- 0.147 # 0.147 for Sweden
  }
  if (countryAbbrev == "UK"){
    sigmaGuess <- 0.5
    gamma_EGuess <- 0.147 # 0.147 for Sweden
    betaGuess <- 0.9
    lambda_LGuess <- 0.0
    lambda_EGuess <- 0.0
    lower = list(sigma=-Inf, gamma_E=0.0, beta=-Inf, lambda_L=-0.3, lambda_E=-0.3)
    upper = list(sigma= Inf, gamma_E=1.0, beta= Inf, lambda_L= 0.3, lambda_E= 0.3)
  }
  if (countryAbbrev == "CN"){

  }
  # Runs a non-linear least squares fit to the data with constraints
#   modelCES <- nls(iGDP ~ ((1-gamma_E)^(1.0/sigma) * ((exp(lambda_L*iYear))^beta 
#                                                   * iLabor^beta
#                                                   * iCapStk^(1-beta) )^((sigma-1.0)/sigma) 
#                           + gamma_E^(1.0/sigma)*(exp(lambda_E*iYear) 
#                                                   * iEToFit)^((sigma-1.0)/sigma))^(sigma/(sigma-1.0)), 
#                   data=dataTable,
#                   start=list(sigma=sigmaGuess, gamma_E=gamma_EGuess, beta=betaGuess, 
#                              lambda_L=lambda_LGuess, lambda_E=lambda_EGuess),
#                   control=nls.control(maxiter=10000, tol=1e-05, minFactor=1/1024, 
#                                       printEval=TRUE, warnOnly=TRUE),
#                   algorithm = "port",
#                   lower = lower,
#                   upper = upper
#                   )
  model <- iGDP ~ ((1-gamma_E)^(1.0/sigma) * ((exp(lambda_L*iYear))^beta 
                                                  * iLabor^beta
                                                  * iCapStk^(1-beta) )^((sigma-1.0)/sigma) 
                          + gamma_E^(1.0/sigma)*(exp(lambda_E*iYear) 
                                                  * iEToFit)^((sigma-1.0)/sigma))^(sigma/(sigma-1.0))
  modelCES <- nlxb(model,
                   data=dataTable,
                   start=list(sigma=sigmaGuess, gamma_E=gamma_EGuess, beta=betaGuess, 
                              lambda_L=lambda_LGuess, lambda_E=lambda_EGuess)
                   )
print(modelCES)
  return(modelCES)
}
print("US Q")
print(summary(cesModel("US", "Q")))

@@

The CES model has the form

\begin{equation} \label{eq:CES_model}
  y = \left\{ (1-\gamma_E)^{1/\sigma} \left[ k^{\alpha} \left( \mathrm{e}^{\lambda_L(t-t_0)} l \right)^{\beta} \right]^{\frac{\sigma-1}{\sigma}}   
      + (\gamma_E)^{1/\sigma} \left[ \mathrm{e}^{\lambda_E(t-t_0)} e \right]^{\frac{\sigma - 1}{\sigma}}   \right\}
      ^{\frac{\sigma}{\sigma - 1}},
\end{equation}

\noindent where $e$ is any of $q$, $x$, or $u$; $\sigma$, $\gamma_E$, $\beta$, $\lambda_L$, and $\lambda_E$ are fitting parameters; and $\mathrm{e}$ is the base of the natural logarithm. Ths sum $\alpha + \beta$ is constrained to unity.

>>>>

\usepackage{soul} %Provides strikethrough text
\usepackage{float} %Allows precise positioning of tables and figures within the text
\usepackage{amsmath} %Allows \begin{equation*} for unnumbered equations

@@

<<setup, echo=FALSE, message=FALSE>>=
require(knitr)
require(lattice)
require(latticeExtra)
require(ggplot2)
require(car)
require(mosaic)
require(xtable)
require(nlmrt)
# tikz allows use of LaTeX formatting and font in graphs. Allows for a consistent look across the paper.
# See http://r-forge.r-project.org/R/?group_id=440 for instructions on installing tikzDevice.
require(tikzDevice) 
trellis.par.set(theme=theme.mosaic())
options(width=75)
options(show.signif.stars=TRUE)
opts_chunk$set(
  # ********************************************************************************************************
  eval=FALSE,
  # ********************************************************************************************************
  warning=TRUE, #Tells whether to show warnings in the output.
  dev='tikz',    #Allows LaTeX code in graphical output. E.g., "$y$" for a variable name in the legend for a graph.
  tidy=FALSE,
  comment=NA,
  #Tells whether to cache output from chunks are saved. Cacheing saves time. 
  #However, references to LaTeX tables DO NOT WORK if cacheing is turned on (cache=TRUE).
  cache=FALSE,
  echo=FALSE,    #Tells whether to echo code for all chunks. Set TRUE on an individual chunk to see its code.
  message=FALSE) #Turns off messages for all chunks. Set TRUE on an individual chunk to see it.

@@

<<nlmrt tests, eval=FALSE>>=
# From https://stat.ethz.ch/pipermail/r-help/2012-November/330554.html
Laptop_sale <- c(1405,1863,2027,2669,2938,5275,6595,6943,8621,10905,12420,22400,32380,31600,34900,43163,47838,47592)
CuSale <- Laptop_sale
time <- seq_along(Laptop_sale)
Bass.Model <- Laptop_sale ~ m * ((p + q)^2/p) * (exp(-(p + q) * time)/((q /p) * exp(-(p + q) * time) + 1)^2)
# Bass.Fit <- nls(formula = Bass.Model, start = c(p = 0.03, q = 0.4, m = max(CuSale)), trace = FALSE)
# Bass.Fit
library(nlmrt)
xdata<-data.frame(Laptop_sale, time)
Bass.Fitn <- nlxb(formula = Bass.Model, start = c(p = 0.03, q = 0.4, m = max(CuSale)), data=xdata,trace=TRUE)
print("")
print("After nlxb fit:")
Bass.Fitn
print(summary(Bass.Fitn))

Bass.Fitn <- wrapnls(formula = Bass.Model, start = c(p = 0.03, q = 0.4, m = max(CuSale)), data=xdata,trace=FALSE)
print("")
print("After wrapnls fit:")
print(summary(Bass.Fitn))
@

@@

<<single-factor functions, eval=TRUE>>=
####################
# Returns an nls single-factor model for the country and factor specified.
# factor should be one of "K", "L", "Q", "X", or "U".
##
singleFactorModel <- function(countryAbbrev, factor){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  # We'll change the name of the desired column to "f"
  dataTable <- replaceColName(dataTable, factor, "f")
  # Now do the fit.
  lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
  # Some economies need different guesses
  if (countryAbbrev == "IR" && factor == "X"){
    mGuess <- 0.7
  } else {
    # This works for every other combination of country and factor.
    mGuess <- 0.5
  }
  start <- list(lambda=lambdaGuess, m=mGuess)
  # Runs a non-linear least squares fit to the data. We've replaced beta with 1-alpha for simplicity.
  model <- iGDP ~ exp(lambda*iYear) * f^m
  modelSF <- nls(formula=model, data=dataTable, start = start)
#   model <- iGDP ~ exp(lambda*iYear) * f^m
#   modelSF <- wrapnls(formula=model, data=dataTable, start=list(lambda=lambdaGuess, m=mGuess))
  return(modelSF)
}

@@

\noindent to historical data where the factor of production ($f$) is any of $k$, $l$, $q$, $x$, or $u$, and the exponent ($m$) is $\alpha$ for $k$, $\beta$ for $l$, or $\gamma$ for $q$, $x$, and $u$.

@@

<<cobb-douglas functions, eval=TRUE>>=
####################
# Returns an nls Cobb-Douglas model for the country specified
# Give an energyType ("Q", "X", or "U") if you want to include an energy term
##
cobbDouglasModel <- function(countryAbbrev, energyType){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  #Run the non-linear least squares fit to the data based on whether we have an energy type or not.
  if (missing(energyType)){
    # No energy term desired in the Cobb-Douglas equation.
    # Establish guess values for alpha and lambda.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.7 #0.7 gives good results for all countries.  
    start <- list(lambda=lambdaGuess, alpha=alphaGuess)
    # Runs a non-linear least squares fit to the data. We've replaced beta with 1-alpha for simplicity.
    model <- iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^(1 - alpha)
    modelCD <- nls(formula=model, data=dataTable, start = start,
                   #Include the next 3 lines to fit with constraints.
                   #algorithm = "port",
                   #lower = list(lambda=-Inf, alpha=0), #constraints on lambda and alpha
                   #upper = list(lambda=Inf, alpha=1)
                   )
    return(modelCD)
  } else {
    # We need to do the Cobb-Douglas fit with the desired energy data.
    # To achieve the correct fit, we'll change the name of the desired column
    # to "iEToFit" and use "iEToFit" in the nls function. 
    dataTable <- replaceColName(dataTable, energyType, "iEToFit")
    # Reparameterize to ensure that we meet the constraints:
    # * alpha + beta + gamma = 1.0.
    # * alpha, beta, and gamma are all between 0.0 and 1.0.
    # To do this, we reparameterize as
    # * 0 < a < 1
    # * 0 < b < 1
    # * alpha = min(a, b)
    # * beta = b - a
    # * gamma = 1 - max(a, b)
    # Establish guess values for lambda, alpha, and beta.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.9
    betaGuess <- 1.0 - alphaGuess
    aGuess <- alphaGuess
    bGuess <- alphaGuess + betaGuess
    start <- list(lambda=lambdaGuess, a=aGuess, b=bGuess)
    # Now actually do the fit, using the column name "iEToFit".
    model <- iGDP ~ exp(lambda*iYear) * iCapStk^min(a,b) * iLabor^abs(b-a) * iEToFit^(1.0 - max(a,b))
    modelCDe <- nls(formula=model, data = dataTable, start = start,
                    control = nls.control(maxiter = 200, 
                                          tol = 1e-05, 
                                          minFactor = 1/1024, 
                                          printEval=FALSE, #Tells whether to print details of curve fit process.
                                          warnOnly=TRUE),
                    #Include the next 3 lines to fit with constraints.
                    #algorithm = "port",
                    #lower = list(lambda=-Inf, a=0, b=0),
                    #upper = list(lambda= Inf, a=1, b=1)
                    )
    return(modelCDe)
  }
}

@@

<<ces functions, eval=TRUE>>=
####################
# Returns an nls ces model for the country and energyType specified.
# energyType should be one of Q", "X", or "U".
##
cesModel <- function(countryAbbrev, energyType){
  # Load the data that we need.
  dataTable <- loadData(countryAbbrev)
  # We need to do the CES fit with the desired energyType.
  # To achieve the correct fit, we'll change the name of the desired column
  # to "iEToFit" and use "iEToFit" in the nls function.
  dataTable <- replaceColName(dataTable, energyType, "iEToFit")
  # Establish default guesses for sigma, gamma_E, beta, lambda_L and lambda_E.
  start <- list(sigma=0.9, gamma_E=0.01, beta=0.7, lambda_L=0.011, lambda_E=0.018) #values for Sweden
  # Establish bounds
  lower = list(sigma=-Inf, gamma_E=0.0, beta=-Inf, lambda_L=-Inf, lambda_E=-Inf)
  upper = list(sigma= Inf, gamma_E=1.0, beta= Inf, lambda_L= Inf, lambda_E= Inf)
  # Now some country-specific values.
  if (countryAbbrev == "US"){
    start <- list(sigma=0.95, gamma_E=0.1, beta=0.7, lambda_L=0.015, lambda_E=0.015) #values for Sweden
  }
  if (countryAbbrev == "UK"){
    start <- list(sigma=0.95, gamma_E=0.2, beta=0.7, lambda_L=0.02, lambda_E=0.02) #values for Sweden
    lower = list(sigma=-Inf, gamma_E=0.0, beta=-Inf, lambda_L=-0.3, lambda_E=-0.3)
    upper = list(sigma= Inf, gamma_E=1.0, beta= Inf, lambda_L= 0.3, lambda_E= 0.3)
  }
  if (countryAbbrev == "CN"){

  }
  # Runs a non-linear least squares fit to the data with constraints
#   model <- iGDP ~ ((1-gamma_E)^(1.0/sigma) * ((exp(lambda_L*iYear))^beta 
#                                                   * iLabor^beta
#                                                   * iCapStk^(1-beta) )^((sigma-1.0)/sigma) 
#                   + gamma_E^(1.0/sigma)*(exp(lambda_E*iYear) * iEToFit)^((sigma-1.0)/sigma))^(sigma/(sigma-1.0))
  model <- iGDP ~ ((1-gamma_E) * ((exp(lambda_L*iYear))^beta * iLabor^beta * iCapStk^(1-beta) )^((sigma-1.0)/sigma) 
                  + gamma_E * (exp(lambda_E*iYear) * iEToFit)^((sigma-1.0)/sigma))^(sigma/(sigma-1.0))
  modelCES <- nls(formula=model, data=dataTable, start=start,
                  control=nls.control(maxiter=10000, tol=1e-05, minFactor=1/1024, 
                                      printEval=TRUE, warnOnly=TRUE),
                  algorithm = "port",
                  lower = lower,
                  upper = upper,
                  trace=TRUE
                  )
#   modelCES <- nlxb(model,
#                    data=dataTable,
#                    start=list(sigma=sigmaGuess, gamma_E=gamma_EGuess, beta=betaGuess, 
#                               lambda_L=lambda_LGuess, lambda_E=lambda_EGuess),
#                    trace=TRUE
#                   )
  return(modelCES)
}
print("US Q")
print(summary(cesModel("US", "Q")))

@@

\citet{Stern:2012ey} developed a CES model of the form

\begin{equation} \label{eq:CES_model}
  y = \left\{ (1-\gamma_E) \left[ k^{\alpha} \left( \mathrm{e}^{\lambda_L(t-t_0)} l \right)^{\beta} \right]^{\frac{\sigma-1}{\sigma}}   
      + (\gamma_E) \left[ \mathrm{e}^{\lambda_E(t-t_0)} e \right]^{\frac{\sigma - 1}{\sigma}}   \right\}
      ^{\frac{\sigma}{\sigma - 1}},
\end{equation}

\noindent where $e$ is any of $q$, $x$, or $u$; $\sigma$, $\gamma_E$, $\beta$, $\lambda_L$, and $\lambda_E$ are fitting parameters; $\mathrm{e}$ is the base of the natural logarithm; and the sum $\alpha + \beta$ is constrained to unity.\footnote{The CES equation in \citet{Stern:2012ey} appears as

\begin{equation*} \label{eq:Stern_CES_model}
  y = \left\{ (1-\gamma_E)^{1/\sigma} \left[ k^{\alpha} \left( \mathrm{e}^{\lambda_L(t-t_0)} l \right)^{\beta} \right]^{\frac{\sigma-1}{\sigma}}   
      + (\gamma_E)^{1/\sigma} \left[ \mathrm{e}^{\lambda_E(t-t_0)} e \right]^{\frac{\sigma - 1}{\sigma}}   \right\}
      ^{\frac{\sigma}{\sigma - 1}},
\end{equation*}

but the $1/\sigma$ exponent on the $(1-\gamma_E)$ and $\gamma_E$ terms seems erroneous due to $y(t_\mathrm{0}) \neq 1$.}
